Icicle: write once, run once

### 2016-05-26 2016fhpc/section/02-Source.tex
some minor changes and remove GVN stuff from related

% ---------------------------------------------------------
\subsection{The Stage Restriction}
To ensure that Icicle queries can be evaluated in a single pass, we use a modal type system inspired by staged computation~\cite{davies2001modal}. We use two modalities, @Element@ and @Aggregate@. Values of type @Element@~$\tau$ are taken from input stream on a per-element basis, whereas values of type @Aggregate@~$\tau$ are only available once the entire stream has been consumed. In the expression @(filter (key == k) of mean value)@, the variable @key@ has type @Element Text@ while @k@ has type @Aggregate Text@. Attempting to compile this query in Icicle will produce a type error complaining that elements cannot be compared with aggregates.

@@

Suppose we apply a query large table and record the result. Tomorrow morning we receive more data and add it to the table. We would like to recompute the result of this query without needing to process all data from the start of the table. We can do this by remembering the values of all intermediate aggregates that are computed in the query, and updating them as new data arrives. In the @avg@ example from \S2 these aggregates are @k@ and @avgs@.

@@

% ---------------------------------------------------------
\subsection{Bounded buffer restriction}
\label{s:IcicleSource:bounded}
Icicle queries are intended to process tables of arbitrary size that may not fit in memory. Due to this, each query must execute without requiring buffer space proportional to the size of the input. For example, here is a simple Haskell expression which cannot be executed without reserving a buffer at least the same size as the input:
\begin{code}
      zip (filter (> 0) xs) (filter (< 0) xs)
\end{code}

Suppose @(xs :: Stream Int)@. The expression takes the input stream @xs@ and selects the elements greater than zero, and pairs them with elements that are less than zero. This expression requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then the positive values must be retained before any output can be produced.

The language constructs of Icicle are designed so that queries that would require unbounded buffering cannot be written, with a one major cavet that we will discuss in a moment. In Icicle, the stream being processed (such as @xs@ above) is implicit in the defined query, and is never named explicitly. Constructs such as @filter@ and @fold@ do not take the name of the stream as argument, but instead operate on stream defined in the context. In all the example Icicle queries so far, the name of the table / stream is not mentioned in the body of the query, and serves as a comment only. Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query. In the expression @filter p in mean value@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.

>>>>

% ---------------------------------------------------------
\subsection{The Stage Restriction}
To ensure that Icicle queries can be evaluated in a single pass, we use a modal type system inspired by staged computation~\cite{davies2001modal}. We use two modalities, @Element@ and @Aggregate@. Values of type @Element@~$\tau$ are taken from input stream on a per-element basis, whereas values of type @Aggregate@~$\tau$ are only available once the entire stream has been consumed. In the expression @(filter (key == k) of mean value)@, the variable @key@ has type @Element Date@ while @k@ has type @Aggregate Date@. Attempting to compile this query in Icicle will produce a type error complaining that elements cannot be compared with aggregates.

@@

Suppose we apply a query to a large table and record the result. Tomorrow morning we receive more data and add it to the table. We would like to recompute the result of this query without needing to process all data from the start of the table. We can do this by remembering the values of all intermediate aggregates that are computed in the query, and updating them as new data arrives. In the @avg@ example from \S2 these aggregates are @k@ and @avgs@.

@@

% ---------------------------------------------------------
\subsection{Bounded buffer restriction}
\label{s:IcicleSource:bounded}
Icicle queries are intended to process tables of arbitrary size that may not fit in memory. Due to this, each query must execute without requiring buffer space proportional to the size of the input. For example, here is a simple Haskell expression which cannot be executed without reserving a buffer at least the same size as the input:
\begin{code}
    unbounded (xs :: Stream Int)
     = zip (filter (> 0) xs) (filter (< 0) xs)
\end{code}

The expression takes the input stream @xs@ and selects the elements greater than zero, and pairs them with elements that are less than zero. This expression requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then the positive values must be retained before any output can be produced.

The language constructs of Icicle are designed so that queries that would require unbounded buffering cannot be written, with one major caveat that we will discuss in a moment. In Icicle, the stream being processed (such as @xs@ above) is implicit in the defined query, and is never named explicitly. Constructs such as @filter@ and @fold@ do not take the name of the stream as argument, but instead operate on stream defined in the context. In all the example Icicle queries so far, the name of the table / stream is not mentioned in the body of the query, and serves as a comment only. Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query. In the expression @filter p of mean value@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.

### 2016-05-26 2016fhpc/section/XX-Related.tex
some minor changes and remove GVN stuff from related

``most compilers restrict themselves to common subexpression elimination within basic blocks or extended basic blocks''
\cite{debray1992compiler}

Temporal common subexpression elimination in Single Assignment C
allows reuse of expressions computed in the previous iteration\cite{imlig2001loop}.
This is a more general case of the elimination opportunities that arise from loop unrolling.

For example, the program below loops over an array @A@, and stores the product @a@ of some computation on the current element (@f(A[i])@), while performing the same computation over the next element (@f(A[i+1])@) and summing it in @b@.
On successive iterations, the @bx = f(A[i+1])@ computed from last iteration could be reused as the new @ax = f(A[i])@.
\begin{code}
int[] A;
int a = 1;
int b = 0;
while(int i = 0; i != size - 1; i++)  {
  int ax = f(A[i]);
  int bx = f(A[i+1]);
  a += ax;
  b *= bx;
}
\end{code}

I cannot think of a better example than this right now.

@@

Multi-query data analysis\cite{andrade2003efficient} describes a similar problem, where they have multiple queries over the same data that run periodically.
By grouping the multiple queries together, more optimisations can be performed than when treating the queries separately.
This is a lot closer, and performs CSE on expressions like @x = sum(y)@, where @sum@ is a built-in aggregate function.
However, this is only for a small set of built-in aggregates, and does not deal with mutually recursive

>>>>

\TODO{fold this into related}
The Icicle approach to bounded buffering is simpler but less expressive than both existing synchronous data flow languages such as Lucy~\cite{mandel2010lucy}, as well as our prior work on flow fusion~\cite{lippmeier2013data}. These systems perform a clock, or \emph{rate} analysis of the streams being processed. Stream combinators with multiple inputs, such as @zip@, are assigned types that require their stream arguments to have the same rate --- meaning that elements always arrive in lockstep and the combinators themselves do not need to perform their own buffering.

% The most closely related transforms are induction variable elimination\cite{shivers1988control} and global value numbering\cite{rosen1988global}.
% Induction variable elimination finds and removes accumulators that are linear functions of the iteration number.
% This does not deal with non-linear functions such as sum, or mutually recursive accumulators.

% Global value numbering works on an SSA form and is much more general than induction variable elimination.
% Early versions such as Rosen et al\cite{rosen1988global} only worked on extended basic blocks with backedges removed.
% Alpern et al\cite{alpern1988detecting} introduced cases for specific loop backedges, where a fixpoint is performed to find the congruence sets of loops.
% More recently, Gulwani and Necula\cite{gulwani2004polynomial} improved upon this, supporting removal of more expressions, and imposing a polynomial time bound.
% Global value numbering for loops can perform all optimisations here.

% Even better is Nie and Cheng's approach\cite{nie2007efficient}, which does not seem to be more complete or lower complexity than Gulwani and Necula, just lower constant factors.

% Common subexpression elimination is very closely related, but most imperative compilers only perform common subexpression elimination on straight-line computations with no control flow\cite{debray1992compiler}, or on loop invariant expressions\cite{bodik1998complete}.
% Common subexpression elimination for functional languages would be able remove lone expressions, but does not deal with mutually recursive folds.

% Temporal common subexpression elimination in Single Assignment C allows reuse of expressions computed in the previous iteration\cite{imlig2001loop}.
% This is a more general case of the elimination opportunities that arise from loop unrolling.

% For example, the program below loops over an array @A@, and stores the product @a@ of some computation on the current element (@f(A[i])@), while performing the same computation over the next element (@f(A[i+1])@) and summing it in @b@.
% On successive iterations, the @bx = f(A[i+1])@ computed from last iteration could be reused as the new @ax = f(A[i])@.
% \begin{code}
% int[] A;
% int a = 1;
% int b = 0;
% while(int i = 0; i != size - 1; i++)  {
%   int ax = f(A[i]);
%   int bx = f(A[i+1]);
%   a += ax;
%   b *= bx;
% }
% \end{code}
% 
% I cannot think of a better example than this right now.

@@

Multi-query data analysis\cite{andrade2003efficient} describes a similar problem, where they have multiple queries over the same data that run periodically.
By grouping the multiple queries together, more optimisations can be performed than when treating the queries separately.
This is a lot closer, and performs CSE on expressions like @x = sum(y)@, where @sum@ is a built-in aggregate function.
However, this is only for a small set of built-in aggregates, and does not deal with mutually recursive definitions.

% The polyhedral model\cite{benabderrahmane2010polyhedral} analyses data dependencies in loops.
% For a given iteration, the polyhedral model computes the iterations which must be executed before the given iteration.
% By knowing the dependent iterations, one can restructure the loop so that the iterations are executed in a different order, while still performing any dependencies in order.
% For example, a loop iterating over @i@, with loop body @A[i] = A[i-10]@ depends on the tenth previous iteration.
% This loop could be executed in an order like @0, 1, 2,...@, but it could also stride by ten: @0, 10, 20..., 1, 11, 21,...@.
% This allows fusion of any two loops if the intersection of the two loops' iteration spaces is not empty.
% In our case the iteration order is fixed, as the program cannot control the order in which stream elements are received.
% The polyhedral model may still be applicable to streaming computations where one could introduce certain-sized buffers, then process the buffer in parallel.
% In this case the polyhedral model may expose more opportunities for duplicate elimination, but does not itself remove duplicates.
%
